Validation of Visual Statistical Inference, Applied to Linear Models

### 2013-03-05 paper_jasa.tex
actual plot location hidden

\begin {abstract}  
Statistical graphics play a crucial role in exploratory data analysis, model checking and diagnosis. A new method, called the lineup protocol, enables determining the statistical significance of visual findings, which bridges a big gulf between exploratory and inferential statistics.  In this paper inferential methods for statistical graphics is developed further by refining the terminology of visual inference, and framing the lineup protocol in a context that allows direct comparison with conventional tests, in scenarios when a conventional test exists. This framework is used to compare the performance of the lineup protocol against conventional statistical testing in the scenario of fitting linear models. A human subjects experiment is conducted using simulated data to provide controlled conditions. Results suggest that the lineup protocol provides comparable results to the conventional tests, and expectedly out-performs them when data is contaminated, a scenario where assumptions required for performing a conventional test are violated, and there may be some super-individuals who yield better power than the conventional test even in the most difficult tasks.

@@

Statistical graphics nourish the discovery process in data analysis by revealing unexpected things,  finding structure that was not previously anticipated,  or orthogonally by contrasting prevailing hypotheses. The area of graphics is often associated with exploratory data analysis, which was pioneered by \cite{tukey:eda} and, is particularly pertinent in today's data-rich world where discovery during data mining has become an important activity. Graphics are also used in many places where numerical summaries simply do not suffice: model checking, diagnosis, and in the communication of findings.

Statistical graphics research and development is agreeably established now, and has been cemented by several new developments in recent years. Early studies on evaluating how well statistical plots are perceived and read by the human eye \citep{cleveland:1984}, have been repeated and expanded \citep{simkin:1987,spence:1991, heer:2010} with findings supporting the original results. This body of work provides a framework for evaluating new statistical graphics.  In a complementary direction, new research on formalizing statistical graphics with language characteristics makes it easier to abstractly define, compare and contrast data plots. \cite{wilkinson:1999} developed a grammar of graphics that is enhanced by \cite{hadley:2009}. These methods provide a mechanism to abstract the way data is mapped to graphical form. Finally, technology advances make it simple and easy for everyone to draw plots of data, and particularly the existence of software systems, such as R \citep{R}, enable making beautiful data graphics that can be tightly coupled with statistical modeling.

@@

The protocol has only been informally tested until now. In the work described in this paper, the lineup protocol is compared head to head with the equivalent conventional test. Specifically, the lineup is examined in the context of a linear model setting, where we are determining the importance of including a variable in the model. This is not the envisioned environment for the use of the lineup -- actually it is likely the worst case scenario for visual inference. The intended use of lineups is where there is no existing test, and unlikely to be ever any numerical test. The thought is though, that the conventional setting provides a benchmark for how well the lineup protocol works under controlled conditions, and will provide some assurance that they will work in scenarios where there is no benchmark. Testing is done based on a human-subjects experiment using Amazon's Mechanical Turk \citep{turk}, using simulation to provide controlled conditions for assessing lineups. The results are compared with those of the conventional test.

@@

Table \ref{tbl:compare} illustrates the lineup protocol in relation to conventional hypothesis testing. Both start from the same place, the same set of hypotheses. The conventional test statistic is the $t$-statistic, where the parameter estimate is divided by its standard error. In the lineup protocol, the test statistic is a plot of the data. Here, side-by-side boxplots are used, because the variable of interest is categorical and takes just two values. In conventional hypothesis testing the value of the test statistic is compared with all possible values of the sampling distribution, the distribution of the statistic if the null hypothesis is true. If it is extreme on this scale then the null hypothesis is rejected. In contrast in visual inference, the plot of the data is compared with plots of a finite number of samples drawn from the null distribution. If the actual data plot is selected as the most different, then this results in rejection of the null hypothesis.

@@

\noindent where $g(.)$ denotes the {\it logit} link function $g(\pi)=\log(\pi) - \log(1-\pi); 0 \le \pi \le 1$.
$W$ is a design matrix of covariates corresponding to specifics of lineup $\ell$ and subject $i$, and $\delta$ is the vector of corresponding parameters. Covariates could include  demographic information of individuals, such as age, gender, education level etc.,  as well lineup-specific elements, e.g. effect size or difficulty level.
%
$Z_{\ell i}$,  $1 \le i \le K$, $1 \le \ell \le L$,  is a design matrix corresponding to random effects specific to individual $i$ and lineup $\ell$; and 
$\tau$  is a vector of independent normally distributed random variables $\tau_{\ell i}$ with  variance matrix $\sigma_\tau I_{KL \times KL}$. $\tau$ will usually include a component incorporating an individual's ability or skill to evaluate lineups. 
Note that $\tau_{\ell i }$ usually only includes a partial interaction; for a full interaction of subjects' skills and lineup-specific difficulty we would need replicates of the same subject evaluating the same lineup, which in practice is not feasible without loosing independence.

@@

In order to compare the power of conventional and visual tests side-by-side, it is necessary to assume, that we are in the controlled environment of a simulation with tests corresponding to a known parameter value $\theta \in R$ and associated distribution function $F_t$ of the test statistic.

@@

The experiment is designed to study the ability of human subjects to detect the effect of a single categorical variable $X_2$ (corresponding to parameter $\beta_2$) in a two variable ($p=2$) regression model (Equation \ref{multi}). Data is simulated using a range of values of $\beta_2$ or slopes as shown in Table \ref{tbl:experiment_params}, two different sample sizes ($n=100, 300$) and two standard deviations of the error ($\sigma=5, 12$). The range of $\beta_2$ values was chosen so that estimates of the power would produce reasonably continuous power curves, comparable to that calculated for the theoretical conventional test. Values were fixed for other regression parameters, $\beta_0 = 5$,  $\beta_1=15$, and the values for $X_1$ were randomly generated from a Poisson $(\lambda=30)$ distribution, which is almost Gaussian. Data sets were generated with parameter values shown in Table \ref{tbl:experiment_params}. Three replicates at each level were generated. These produced 60 different ``actual data sets'', and thus, 60 different lineups. To produce the lineups, the null model was fit to each actual data set, to obtain residuals and parameter estimates. The actual data plot was generated by making  side-by-side boxplots of the residuals (Table \ref{tbl:stat_multiple}, case 3). The 19 null data sets were generated by simulating from  $N(0, {\hat{\sigma}}^2)$, and  plotted in the same way. The actual data plot was randomly placed among these null data plots to produce the lineup.  \blue {Figure \ref{fig:test_category} is an example of one of these lineups. The actual data plot is number 15 which was generated for $n$=300, $\beta_2$=10 and $\sigma$=12. For this lineup, 15 out of 16 observers picked the actual data plot. }

@@

The number of evaluations required for each lineup to provide reasonable estimates of the power is determined by the variance of the estimated power. Suppose $\gamma$ denotes the conventional test power for each parameter combination shown in Table \ref{tbl:experiment_params}. Since the expected power of visual inference is very close to the power of conventional test (Figure \ref{fig:power_expected}) the proportion of successful evaluations of lineups should give a rough estimate for  $\gamma$. 
For a given proportion $\gamma$ it is desired to have a margin of error (ME) less than or equal to 0.05. Thus we have $ME =1.96 \sqrt{ \gamma(1-\gamma) / n_{\gamma} } \le 0.05$ which gives us the estimation of minimum number of evaluations $$n_{\gamma} \geq \frac{\gamma(1-\gamma)}{(0.05/1.96)^2}.$$  
% \hh{we're not even close to $n_\gamma$ with ten lineups .... we need more in the order of 400. Let's not emphasize this so much.}

@@

The actual data plot is generated by making a scatterplot of $Y$ vs $X_1$ with the least squares regression line overlaid. To produce the null plots in the lineup null data was simulated from $N(X \hat{\beta}, {\hat{\sigma}}^2$) and plotted using the same scatterplot method as the actual data. 
To select 10 lineups for a subject, each combination of sample size ($n$) and error SD ($\sigma$) is given a difficulty value based on the slope ($\beta$) parameters. For the smallest slopes the difficulty is 4 (hardest) and for the largest slopes the difficulty is 0 (easiest). 
Figure \ref{fig:test_continuous} shows an example lineup for this experiment from difficulty level 4.  This lineup is generated using sample size ($n$)=100, slope($\beta$)=1.25 and and error SD($\sigma$)=5. \blue{The actual data plot location is 5. None of the 65 observers picked the actual plot while 46 observers picked plot 18 which has the lowest $p$-value among all the plots in this lineup as we can see in Figure \ref{fig:P-val_log2} for Experiment 2.}

For each combination of sample size and standard deviation, the subject sees a randomly selected lineup of each difficulty level, all the slopes, giving 5 lineups. Another four lineups are chosen from a second choice of sample size and standard deviation, with difficulty levels 0-3, and the last lineup shown was a randomly selected lineup with difficulty 0. The order that the lineups are shown to participants is randomized. Each subject saw a substantial number of difficult lineups.

@@

An example lineup for slope $\beta=0.4$ is shown in Figure \ref{fig:test_contaminated}.  Can you pick which plot is different? \blue{The actual data plot location is 1. Out of 31 observers 13 observers picked the actual plot.}

@@

%{\bf Screening scenarios:}
Amazon Mechanical Turk workers are paid for their efforts, not substantial, but on the scale of the minimum wage of the USA. Some workers will try to maximize their earnings for minimum effort, which can affect the results from the data. For example, some workers may simply randomly pick a plot, without actively examining the plots in the lineup. For the purpose of identifying these participants and cleaning the data, we take action based on a subject's answer to one of the very easy lineups that everybody was exposed to:
if the subject failed to identify the actual data plot, we remove all of this subject's data from the analysis. If the answer on the lineup is correct, we remove this lineup from the analysis, but keep all of the remaining answers. Table \ref{tbl:summary} tabulates the number of subjects, genders and lineups evaluated after applying the data screening procedure.  \blue {A summary of all the data along with cleaned data are added as the supplementary material.}

@@

\begin{figure}[hbtp]
   \centering
       \scalebox{0.60}{\includegraphics{power_mixed_subject_k.pdf}}
       \caption{Subject-specific  power estimated from model \ref{eqn:mixed} are shown. The corresponding power curve for conventional test (dashed line) is shown for comparison. The overall average power curve estimated from the model is shown (light blue).}
       \label{fig:power_mixed_subject}
\end{figure}

@@

In the real setting, where visual inference is to be useful, there will be no conventional tests $p$-values. Assessing the strength of perceived structure is a critical component of visual inference. In experiments 1 and 2, there is a $p$-value associated with the actual data plot in each lineup. As the $p$-value increases the proportion of correct responses falls (Figure \ref{fig:pval_pcorrect}), which is evidence of direct association between proportion of correct responses and conventional test $p$-values. For  $p$-values  larger than 0.15 it is very uncommon for subjects to correctly identify the actual data plot in the lineup.

@@

\begin{figure}[hbtp]
   \centering
       \scalebox{0.70}{\includegraphics{pval_difference.pdf}}
       \caption{Investigating how the strength of structure in the null plots affects the choice of the actual data plot. Difference between the data plot's $p$-value and the  smallest $p$ value of a null plot vs proportion correct. Negative values indicate the $p$-value of the actual data plot are smaller than those of all of the null plots. Close to zero shows a wide range in the proportion correct, suggesting that when at least one null plot has structure almost as strong as the actual data plot, subjects had a difficult time in making their choice.}
       \label{fig:pval_difference}
\end{figure}

@@

When there is no conventional test available, visual testing procedures can be used due to its few assumptions and non-parametric nature. We have successfully used the procedures informally for local projects. It is hoped that this proves to be extremely valuable in data mining applications, and exploratory analyses, where there are no existing gauges of statistical significance. The lineup protocol operates similarly to general tests, such as two-sided, analysis of variance $F$-tests, and Wilks $\Lambda$ for multivariate analysis of variance. If the null hypothesis is rejected, it is generally that we can say that ``there is something there'' but not specifically what it is in the data that triggers the rejection. Follow-up questions on the reasons provide qualitative insight. Multiple comparisons are often done to refine and understand the results of the general conventional testing, and perhaps some similar approaches might be developed for visual inference.

>>>>

\begin {abstract}  
Statistical graphics play a crucial role in exploratory data analysis, model checking and diagnosis. A new method, called the lineup protocol, \red{enables determining} the statistical significance of visual findings, which bridges a big gulf between exploratory and inferential statistics.  In this paper inferential methods for statistical graphics are developed further by refining the terminology of visual inference, and framing the lineup protocol in a context that allows direct comparison with conventional tests, in scenarios when a conventional test exists. This framework is used to compare the performance of the lineup protocol against conventional statistical testing in the scenario of fitting linear models. A human subjects experiment is conducted using simulated data to provide controlled conditions. \blue{Results suggest that the lineup protocol provides comparable results to the conventional tests, and expectedly out-performs them when signal is statistically significant or data is contaminated, a scenario where assumptions required for performing a conventional test are violated, and there may be some super-individuals who yield better power than the conventional test even in the most difficult tasks.}

@@

Statistical graphics nourish the discovery process in data analysis by revealing unexpected things,  finding structure that was not previously anticipated,  or orthogonally by contrasting prevailing hypotheses. The area of graphics is often associated with exploratory data analysis, which was pioneered by \cite{tukey:eda} and is particularly pertinent in today's data-rich world where discovery during data mining has become an important activity. Graphics are also used in many places where numerical summaries simply do not suffice: model checking, diagnosis, and in the communication of findings.

Statistical graphics research and development is \red{agreeably established} now, and has been cemented by several new developments in recent years. Early studies on evaluating how well statistical plots are perceived and read by the human eye \citep{cleveland:1984}, have been repeated and expanded \citep{simkin:1987,spence:1991, heer:2010} with findings supporting the original results. This body of work provides a framework for evaluating new statistical graphics.  In a complementary direction, new research on formalizing statistical graphics with language characteristics makes it easier to abstractly define, compare and contrast data plots. \cite{wilkinson:1999} developed a grammar of graphics that is enhanced by \cite{hadley:2009}. These methods provide a mechanism to abstract the way data is mapped to graphical form. Finally, technology advances make it simple and easy for everyone to draw plots of data, and particularly the existence of software systems, such as R \citep{R}, enable making beautiful data graphics that can be tightly coupled with statistical modeling.

@@

The protocol has only been informally tested until now. In the work described in this paper, the lineup protocol is compared head to head with the equivalent conventional test. Specifically, the lineup is examined in the context of a linear model setting, where we are determining the importance of including a variable in the model. This is not the envisioned environment for the use of the lineup -- actually it is likely the worst case scenario for visual inference. The intended use of lineups is where there is no existing test, \blue{and unlikely ever to be any numerical test}. The thought is though, that the conventional setting provides a benchmark for how well the lineup protocol works under controlled conditions, and will provide some assurance that they will work in scenarios where there is no benchmark. Testing is done based on a human-subjects experiment using Amazon's Mechanical Turk \citep{turk}, using simulation to provide controlled conditions for assessing lineups. The results are compared with those of the conventional test.

@@

\blue{An illustration of the lineup protocol in relation to conventional hypothesis testing is presented in Table \ref{tbl:compare}.} Both methods start from the same place, the same set of hypotheses. The conventional test statistic is the $t$-statistic, where the parameter estimate is divided by its standard error. In the lineup protocol, the test statistic is a plot of the data. Here, side-by-side boxplots are used, because the variable of interest is categorical and takes just two values. In conventional hypothesis testing the value of the test statistic is compared with all possible values of the sampling distribution, the distribution of the statistic if the null hypothesis is true. If it is extreme on this scale then the null hypothesis is rejected. In contrast in visual inference, the plot of the data is compared with plots of a finite number of samples drawn from the null distribution. If the actual data plot is selected as the most different, then this results in rejection of the null hypothesis.

@@

\noindent where $g(.)$ denotes the {\it logit} link function $g(\pi)=\log(\pi) - \log(1-\pi); 0 \le \pi \le 1$.
$W$ is a design matrix of covariates corresponding to specifics of lineup $\ell$ and subject $i$, and $\delta$ is the vector of corresponding parameters. Covariates could include  demographic information of individuals, such as age, gender, education level etc.,  as well lineup-specific elements, e.g. effect size or difficulty level.
%
$Z_{\ell i}$,  $1 \le i \le K$, $1 \le \ell \le L$,  is a design matrix corresponding to random effects specific to individual $i$ and lineup $\ell$; and 
$\tau$  is a vector of independent normally distributed random variables $\tau_{\ell i}$ with  variance matrix $\sigma_\tau I_{KL \times KL}$. $\tau$ will usually include a component incorporating an individual's ability or skill to evaluate lineups. 
Note that $\tau_{\ell i }$ usually only includes a partial interaction; for a full interaction of subjects' skills and lineup-specific difficulty we would need replicates of the same subject evaluating the same lineup, which in practice is not feasible without \red{loosing} independence.

@@

In order to compare the power of conventional and visual tests side-by-side, it is necessary to assume that we are in the controlled environment of a simulation with tests corresponding to a known parameter value $\theta \in R$ and associated distribution function $F_t$ of the test statistic.

@@

The experiment is designed to study the ability of human subjects to detect the effect of a single categorical variable $X_2$ (corresponding to parameter $\beta_2$) in a two variable ($p=2$) regression model (Equation \ref{multi}). Data is simulated using a range of values of $\beta_2$ or slopes as shown in Table \ref{tbl:experiment_params}, two different sample sizes ($n=100, 300$) and two standard deviations of the error ($\sigma=5, 12$). The range of $\beta_2$ values was chosen so that estimates of the power would produce reasonably continuous power curves, comparable to that calculated for the theoretical conventional test. Values were fixed for other regression parameters, $\beta_0 = 5$,  $\beta_1=15$, and the values for $X_1$ were randomly generated from a Poisson $(\lambda=30)$ distribution, which is almost Gaussian. Data sets were generated with parameter values shown in Table \ref{tbl:experiment_params}. Three replicates at each level were generated. These produced 60 different ``actual data sets'', and thus, 60 different lineups. To produce the lineups, the null model was fit to each actual data set, to obtain residuals and parameter estimates. The actual data plot was generated by making  side-by-side boxplots of the residuals (Table \ref{tbl:stat_multiple}, case 3). The 19 null data sets were generated by simulating from  $N(0, {\hat{\sigma}}^2)$, and  plotted in the same way. The actual data plot was randomly placed among these null data plots to produce the lineup.  \blue {Figure \ref{fig:test_category} is an example of one of these lineups. It was generated for $n$=300, $\beta_2$=10 and $\sigma$=12. The actual data plot location is the integer number nearest to $log_e(3269017)$.  For this lineup, 15 out of 16 observers picked the actual data plot. }

@@

The number of evaluations required for each lineup to provide reasonable estimates of the power is determined by the variance of the estimated power. Suppose $\gamma$ denotes the conventional test power for each parameter combination shown in Table \ref{tbl:experiment_params}. Since the expected power of visual inference is very close to the power of conventional test (Figure \ref{fig:power_expected} with $K=1$) the proportion of successful evaluations of lineups should give a rough estimate for  $\gamma$. 
For a given proportion $\gamma$ it is desired to have a margin of error (ME) less than or equal to 0.05. Thus we have $ME =1.96 \sqrt{ \gamma(1-\gamma) / n_{\gamma} } \le 0.05$ which gives us the estimation of minimum number of evaluations $$n_{\gamma} \geq \frac{\gamma(1-\gamma)}{(0.05/1.96)^2}.$$  
% \hh{we're not even close to $n_\gamma$ with ten lineups .... we need more in the order of 400. Let's not emphasize this so much.}

@@

The actual data plot is generated by making a scatterplot of $Y$ vs $X_1$ with the least squares regression line overlaid. To produce the null plots in the lineup null data was simulated from $N(X \hat{\beta}, {\hat{\sigma}}^2$) and plotted using the same scatterplot method as the actual data. 
To select 10 lineups for a subject, each combination of sample size ($n$) and error SD ($\sigma$) is given a difficulty value based on the slope ($\beta$) parameters. For the smallest slopes the difficulty is 4 (hardest) and for the largest slopes the difficulty is 0 (easiest). 
Figure \ref{fig:test_continuous} shows an example lineup for this experiment from difficulty level 4.  This lineup is generated using sample size ($n$)=100, slope($\beta$)=1.25 and and error SD($\sigma$)=5. \blue{The actual data plot location is the integer value nearest to $log_e(148.4132)$. None of the 65 observers picked the actual plot while 46 observers picked plot 18 which has the lowest $p$-value among all the plots in this lineup as we can see in Figure \ref{fig:P-val_log2} for Experiment 2.}

For each combination of sample size and standard deviation, the subject sees a randomly selected lineup of each difficulty level, giving 5 lineups. Another four lineups are chosen from a second choice of sample size and standard deviation, with difficulty levels 0-3, and the last lineup shown was a randomly selected lineup with difficulty 0. The order that the lineups are shown to participants is randomized. Each subject saw a substantial number of difficult lineups.

@@

An example lineup for slope $\beta=0.4$ is shown in Figure \ref{fig:test_contaminated}.  Can you pick which plot is different? \blue{The actual data plot location is the integer number nearest to $log_e (log_e (15.15426))$ and 13 out of 31 observers picked the actual plot.}

@@

%{\bf Screening scenarios:}
Amazon Mechanical Turk workers are paid for their efforts, not substantially, but on the scale of the minimum wage of the USA. Some workers will try to maximize their earnings for minimum effort, which can affect the results from the data. For example, some workers may simply randomly pick a plot, without actively examining the plots in the lineup. For the purpose of identifying these participants and cleaning the data, we take action based on a subject's answer to one of the very easy lineups that everybody was exposed to:
if the subject failed to identify the actual data plot, we remove all of this subject's data from the analysis. If the answer on the lineup is correct, we remove this lineup from the analysis, but keep all of the remaining answers. Table \ref{tbl:summary} tabulates the number of subjects, genders and lineups evaluated after applying the data screening procedure.  \blue {A summary of all the data along with cleaned data are added to the supplementary material.}

@@

\begin{figure}[hbtp]
   \centering
       \scalebox{0.60}{\includegraphics{power_mixed_subject_k.pdf}}
       \caption{Subject-specific  power estimated from model \ref{eqn:mixed} for $K=5$. The corresponding power curve for conventional test (dashed line) is shown for comparison. The overall average power curve estimated from the model is shown (light blue).}
       \label{fig:power_mixed_subject}
\end{figure}

@@

In the real setting, where visual inference is to be useful, there will be no conventional test $p$-values. Assessing the strength of perceived structure is a critical component of visual inference. In experiments 1 and 2, there is a $p$-value associated with the actual data plot in each lineup. As the $p$-value increases the proportion of correct responses falls (Figure \ref{fig:pval_pcorrect}), which is evidence of direct association between proportion of correct responses and conventional test $p$-values. For  $p$-values  larger than 0.15 it is very uncommon for subjects to correctly identify the actual data plot in the lineup.

@@

\begin{figure}[hbtp]
   \centering
       \scalebox{0.70}{\includegraphics{pval_difference.pdf}}
       \caption{Scatter plot of difference between the data plot's $p$-value and the  smallest $p$-value of the null plots vs proportion correct. Negative differences indicate the $p$-value of the actual data plot are smaller than those of all of the null plots. Difference close to zero shows a wide range in the proportion correct, suggesting that when at least one null plot has structure almost as strong as the actual data plot, subjects had a difficult time in making their choice.}
       \label{fig:pval_difference}
\end{figure}

@@

When there is no conventional test available, visual testing procedure can be used due to its few assumptions and non-parametric nature. We have successfully used the procedure informally for local projects. It is hoped that this proves to be extremely valuable in data mining applications, and exploratory analyses, where there are no existing gauges of statistical significance. The lineup protocol operates similarly to general tests, such as two-sided, analysis of variance $F$-tests, and Wilks $\Lambda$ for multivariate analysis of variance. \blue{If the null hypothesis is rejected, generally we can say that ``there is something there'' but not specifically what it is in the data that triggers the rejection.} Follow-up questions on the reasons provide qualitative insight. Multiple comparisons are often done to refine and understand the results of the general conventional testing, and perhaps some similar approaches might be developed for visual inference.
